State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-16 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
Minor changes in sections 5 and 5.1

%The method is appropriate for state estimation of stochastic nonlinear dynamical systems. 
\section{Results}\label{ResultsSection} The full neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. Note, the parameters for the spatial and temporal aspects of the model can be considered arbitrary when demonstrating the estimation methodology. This is due to the fact that the temporal aspects scale with the sampling rate and the length of the times series, and the spatial aspects scale with the size of the neural field and the space step used for numerical simulation. Figure~\ref{fig:experimental design}(a) shows example data from the sensors of the model and Figure~\ref{fig:experimental design}(b) shows the power spectral density (PSD) of the simulated time-series. The PSD shows the typical $1/f^2$ characteristics of intracranial EEG.
\begin{table}\footnotesize
\centering
\begin{tabular}{ccl}
	\hline\hline Symbol & Value & Units \\
	\hline\hline
	$\tau$ & 0.01 & s$^{-1}$\\
	$f_{max}$ & 10 & spikes/s \\
	$\varsigma$ & 0.8 & spike/mV\\
	$v_0$ & 2 & mV\\
	$\boldsymbol{\theta}$ & $\left[\begin{array}{ccc}
	10 &-8 &0.5
	\end{array}
	\right]^{\top}$ & mV/spike\\
	$\sigma_{\psi_{1}}, \sigma_{\psi_{2}}, \sigma_{\psi_{3}}$ & 1.8, 2.4, 6 & mm\\
	$\sigma_{m},\Delta_{y}$&0.9, 1.5&mm\\
	$\Sigma_{\epsilon}$ &$0.1 \times I_{N} $& mm$^2$ \\
	$\sigma_{\gamma}$&1.3&mm\\
	$\Delta$ & 0.5 & mm \\
	$T_s$ & 1 & ms \\ 
\end{tabular}
\caption{Parameters for the neural field model that are used to generate synthetic data.}
\label{tab:Model Parameters}
\end{table}

The sensors were spaced in a regular square 14 $\times$ 14 grid, the sensor width $\sigma^2_m$ was $0.81$, which provides a width of $1.5$~mm at half the maximum amplitude of the basis function, with spacing $\Delta_y=1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical from neurophysiological recordings. 
\begin{figure}
   	\begin{center}
    		\subfigure[][]{\includegraphics{./Graph/ExperimentFigureEEGPy.pdf}} \\
   		\subfigure[][]{\includegraphics{./Graph/ExperimentFigurePSDPy.pdf}} \\
   	\end{center}
   	\caption{\textbf{a.} Example of data generated by the full neural field model from the first 5 sensors. \textbf{b.} Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of intracranial EEG.} 
\label{fig:experimental design}
\end{figure}
\subsection{Spatial Frequency Analysis} 
Using the model selection techniques described in Section~\ref{SpectralAnalysisSection}, the spatial frequency of the field was used to confirm that the spacing and width of the sensors was adequate to capture the significant dynamics of the field. Figure~\ref{fig:FieldFFT} shows the spatial frequency of the neural field. The cut-off frequency, $\nu_c$, was taken to be 0.24~Hz ($-3$~dB point). From equation~\ref{eq:MinimumSensorDistance}, this yielded a minimum sensor separation of 2.05~mm. This confirms the separation, $\Delta_{y}$, of 1.5~mm was sufficient to prevent problems associated with spatial aliasing.
\begin{figure}
\centering
\subfigure[][]{\label{fig:FieldFFT}\includegraphics{./Graph/FFTTrueField.pdf}}
\subfigure[][]{\label{fig:EstimatedFieldFFT}\includegraphics{./Graph/FFTEstimatedField.pdf}}\\
\caption{Power in dB as a function of spatial frequency of the neural field. \textbf{a.} The average (over time) power of the spatial frequency of the neural field. \textbf{b.} The average (over time) power of the spatial frequency of the reconstructed neural field from the basis function decomposition. The reconstructed field shows a narrower bandwidth as a result of the band limiting properties of the sensors and basis functions.}
\label{fig:FFTTrueEstimate}
\end{figure}
The spatial frequency from the observations is shown in Figure~\ref{fig:ObservationFreqAnal}. The cut-off frequency, $\nu_{cy}$, was taken to be $0.19$~Hz ($-3$~dB point). The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:SensorFunctionFreqResponse}. Using an oversampling parameter, $\rho_{\phi}=1$, the minimum distance between basis functions was calculated to be $2.63$ to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). The distance between basis functions, $\Delta_{\phi}$, was set to $2.5$, which is sufficient to capture the significant dynamics according to our criteria.

Given the basis function separation, the basis function width was set to $\sigma_{\phi}^2=2.5$~mm. This facilitated a smooth representation of the field by the basis function decomposition given the basis function separation and provided low-pass spatial filtering of the field. From equation~\ref{eq:WidthCutOffRelationship}, this allowed the basis function decomposition to represent spatial frequencies up to approximately $0.12$~Hz. The spatial frequency response of the basis functions can be seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
\begin{figure} 
	\begin{center}
		\subfigure[][]{\label{fig:ObservationFreqAnal}\includegraphics{./Graph/ObservationFrequencyResponse.pdf}}\\	
		\subfigure[][]{\label{fig:SensorFunctionFreqResponse}\includegraphics{./Graph/SensorChar.pdf}}
		\subfigure[][]{\label{fig:BasisFunctionFreqResponse}\includegraphics{./Graph/BasisChar.pdf}}
	\end{center}
	\caption{Spatial frequency analysis. \textbf{a.} Two-dimensional spatial frequency in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the 1-dimensional representation of the spectral characteristics. \textbf{b.} The spatial frequency response of a 1-dimensional model sensor. The dashed line shows the cut-off frequency (-3 dB point), indicating a wider bandwidth than the field basis function. \textbf{c.} The spatial frequency response of a 1-dimensional neural field basis function. The dashed line shows the cut-off frequency (-3 dB point).} 
\end{figure}

>>>>

%The method is appropriate for state estimation of stochastic nonlinear dynamical systems. 
\section{Results}\label{ResultsSection} The full neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. Note that, the parameters for the spatial and temporal aspects of the model can be considered arbitrary when demonstrating the estimation methodology. This is due to the fact that the temporal aspects scale with the sampling rate and the length of the times series, and the spatial aspects scale with the size of the neural field and the space step used for numerical simulation. Figure~\ref{fig:experimental design}(a) shows example data from the sensors of the model and Figure~\ref{fig:experimental design}(b) shows the power spectral density (PSD) of the simulated time-series. The PSD shows the typical $1/f^2$ characteristics of intracranial EEG.
\begin{table}\footnotesize
\centering
\begin{tabular}{ccl}
	\hline\hline Symbol & Value & Units \\
	\hline\hline
	$\tau$ & 0.01 & s$^{-1}$\\
	$f_{max}$ & 10 & spikes/s \\
	$\varsigma$ & 0.8 & spike/mV\\
	$v_0$ & 2 & mV\\
	$\boldsymbol{\theta}$ & $\left[\begin{array}{ccc}
	10 &-8 &0.5
	\end{array}
	\right]^{\top}$ & mV/spike\\
	$\sigma_{\psi_{1}}, \sigma_{\psi_{2}}, \sigma_{\psi_{3}}$ & 1.8, 2.4, 6 & mm\\
	$\sigma_{m},\Delta_{y}$&0.9, 1.5&mm\\
	$\Sigma_{\epsilon}$ &$0.1 \times I_{N} $& mm$^2$ \\
	$\sigma_{\gamma}$&1.3&mm\\
	$\Delta$ & 0.5 & mm \\
	$T_s$ & 1 & ms \\ 
\end{tabular}
\caption{Parameters for the neural field model that are used to generate synthetic data.}
\label{tab:Model Parameters}
\end{table}

The sensors were spaced on a regular square 14 $\times$ 14 grid, the sensor width parameter $\sigma^2_m$ was $0.81$, which provides a width of $1.5$~mm at half the maximum amplitude of the basis function, with spacing $\Delta_y=1.5$~mm. This arrangement modelled the recording as having some cross talk, which is typical for neurophysiological recordings. 
\begin{figure}
   	\begin{center}
    		\subfigure[][]{\includegraphics{./Graph/ExperimentFigureEEGPy.pdf}} \\
   		\subfigure[][]{\includegraphics{./Graph/ExperimentFigurePSDPy.pdf}} \\
   	\end{center}
   	\caption{\textbf{a.} Example of data generated by the full neural field model from the first 5 sensors. \textbf{b.} Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of intracranial EEG.} 
\label{fig:experimental design}
\end{figure}
\subsection{Spatial Frequency Analysis} 
Using the model selection techniques described in Section~\ref{SpectralAnalysisSection}, the spatial frequency of the field was used to confirm that the spacing and width of the sensors was adequate to capture the significant dynamics of the field. Figure~\ref{fig:FieldFFT} shows the spatial frequency of the neural field. The cut-off frequency, $\nu_c$, was taken to be 0.24~Hz ($-3$~dB point). From equation~\ref{eq:MinimumSensorDistance}, this yielded a minimum sensor separation of 2.05~mm. This confirms that a separation, $\Delta_{y}$, of 1.5~mm was sufficient to prevent problems associated with spatial aliasing.
\begin{figure}
\centering
\subfigure[][]{\label{fig:FieldFFT}\includegraphics{./Graph/FFTTrueField.pdf}}
\subfigure[][]{\label{fig:EstimatedFieldFFT}\includegraphics{./Graph/FFTEstimatedField.pdf}}\\
\caption{Power in dB as a function of spatial frequency of the neural field. \textbf{a.} The average (over time) power of the spatial frequency of the neural field. \textbf{b.} The average (over time) power of the spatial frequency of the reconstructed neural field from the basis function decomposition. The reconstructed field shows a narrower bandwidth as a result of the band limiting properties of the sensors and basis functions.}
\label{fig:FFTTrueEstimate}
\end{figure}
The spatial frequency of the observed field is shown in Figure~\ref{fig:ObservationFreqAnal}. The cut-off frequency, $\nu_{cy}$, was taken to be $0.19$~Hz ($-3$~dB point). The observation cut-off frequency is lower than the field cut-off due to the low-pass filtering effect of the sensors. The spatial frequency response of the sensors can be seen in Figure~\ref{fig:SensorFunctionFreqResponse}. Using an oversampling parameter, $\rho_{\phi}=1$, the minimum distance between basis functions was calculated to be $2.63$~mm to satisfy Shannon's sampling criterion (from equation~\ref{eq:BasisFunctionSeparation}). The distance between basis functions, $\Delta_{\phi}$, was set to $2.5$~mm, which is sufficient to capture the significant dynamics according to our criteria.

Given the basis function separation, the basis function width parameter was set to $\sigma_{\phi}^2=2.5$~mm. This allowed a smooth representation of the field by the basis function reconstruction up to a spatial frequency as given by equation~\ref{eq:WidthCutOffRelationship} of approximately $0.12$~Hz. This is due to the spatial frequency response of the basis functions as seen in Figure~\ref{fig:BasisFunctionFreqResponse}.
\begin{figure} 
	\begin{center}
		\subfigure[][]{\label{fig:ObservationFreqAnal}\includegraphics{./Graph/ObservationFrequencyResponse.pdf}}\\	
		\subfigure[][]{\label{fig:SensorFunctionFreqResponse}\includegraphics{./Graph/SensorChar.pdf}}
		\subfigure[][]{\label{fig:BasisFunctionFreqResponse}\includegraphics{./Graph/BasisChar.pdf}}
	\end{center}
	\caption{Spatial frequency analysis. \textbf{a.} Two-dimensional spatial frequency in power dB, averaged over 400 time steps) of the observed field. The symmetry of the sensors and field basis functions allow for the 1-dimensional representation of the spectral characteristics. \textbf{b.} The spatial frequency response of a 1-dimensional model sensor. The dashed line shows the cut-off frequency (-3 dB point), indicating a wider bandwidth than the field basis function. \textbf{c.} The spatial frequency response of a 1-dimensional neural field basis function. The dashed line shows the cut-off frequency (-3 dB point).} 
\end{figure}
